Algorithmic detection of conserved quantities for finite-difference schemes\thanks{Supported by King Abdullah University of Science and Technology (KAUST) baseline funds and KAUST OSR-CRG2021-4674.}

### 2023-05-07 Papers/Paper CASC format/paper.tex
paper changes

\documentclass[manuscript]{acmart}

\renewcommand{\labelenumi}{\theenumi.}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage[latin1]{inputenc}
\usepackage{hyperref}
\usepackage{dsfont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{ifthen}
\usepackage{mmacells}
\usepackage[invisible]{MLaTex}

%\usepackage[displaymath,mathlines]{lineno}
%\linenumbers
%\pagewiselinenumbers
%\modulolinenumbers[1]

%permite que formulas
%se dividam por duas
%paginas diferentes
\allowdisplaybreaks

%[hypertex]

%\usepackage{showkeys}
%\usepackage[norefs]{refcheck}

\makeindex

%\usepackage{curves}

%\newcommand{\cal}{\mathcal}
\newcommand{\argmin}{\operatorname{argmin}}
\newcommand{\sech}{\operatorname{sech}}
\newcommand{\intr}{\operatorname{int}}
\newcommand{\arccosh}{\operatorname{arccosh}}
\newcommand{\arcsinh}{\operatorname{arcsinh}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\spn}{\operatorname{span}}
\newcommand{\arccotg}{\operatorname{arccotg}}
\newcommand{\inter}{\operatorname{int}}
\newcommand{\var}{\operatorname{Var}}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\car}{\operatorname{car}}
\newcommand{\cl}{\operatorname{cl}}
\newcommand{\co}{\operatorname{co}}
\newcommand{\arctanh}{\operatorname{arctanh}}
\newcommand{\esssup}{\operatorname{esssup}}
\newcommand{\essinf}{\operatorname{essinf}}
\renewcommand{\div}{\operatorname{div}}
\newcommand{\grad}{\operatorname{grad}}

\newcommand{\Rr}{{\mathbb{R}}}
\newcommand{\Cc}{{\mathbb{C}}}
\newcommand{\Nn}{{\mathbb{N}}}
\newcommand{\Zz}{{\mathbb{Z}}}
\newcommand{\Xx}{{\mathbb{X}}}
\newcommand{\Qq}{{\mathbb{Q}}}
\newcommand{\Tt}{{\mathbb{T}}}
\newcommand{\1}{{\chi}}

\newcommand{\Hh}{{\overline{H}}}
\newcommand{\ot}{{\overline{t}}}
\newcommand{\ox}{{\overline{x}}}
\newcommand{\oy}{{\overline{y}}}
\newcommand{\Ll}{{\mathcal{L}}}

\newcommand{\Aa}{{\mathcal{A}}}
\newcommand{\Cg}{{\mathcal{C}}}
\newcommand{\Ff}{{\mathcal{F}}}
\newcommand{\Gg}{{\mathcal{G}}}
\newcommand{\Ii}{{\mathcal{I}}}
\newcommand{\Dd}{{\mathcal{D}}}
\newcommand{\Uu}{{\mathcal{U}}}
\newcommand{\Vv}{{\mathcal{V}}}
\newcommand{\Ww}{{\mathcal{W}}}
\newcommand{\Mm}{{\mathcal{M}}}

\newcommand{\sfd}{{$\sigma$-field }}
\newcommand{\sfs}{{$\sigma$-fields }}
\newcommand{\cqd}{\hfill $\blacksquare$}
\newcommand{\cqda}{\hfill $\rule{1.5mm}{1.5mm}$}
\newcommand{\cqdb}{\hfill $\rule{0.75mm}{0.75mm}$}
\newcommand{\ee}{\hfill $\blacktriangleleft$}
\newcommand{\eef}{\quad\blacktriangleleft}
\newcommand{\bx}{{\bf x}}
\newcommand{\bu}{{\bf u}}
\newcommand{\bdx}{\dot{\bf x}}
\newcommand{\bp}{{\bf p}}
\newcommand{\bdp}{\dot{\bf p}}
\newcommand{\bX}{{\bf X}}
\newcommand{\bdX}{\dot{\bf X}}
\newcommand{\bP}{{\bf P}}
\newcommand{\bdP}{\dot{\bf P}}
\newcommand{\by}{{\bf y}}
\newcommand{\bdy}{\dot{\bf y}}
\newcommand{\bv}{{\bf v}}
\newcommand{\bdv}{\dot{\bf v}}
\newcommand{\bw}{{\bf w}}
\newcommand{\bdw}{\dot{\bf w}}

\newcommand{\Pf}{{\noindent \sc Proof. \ }}
\newcommand{\Rm}{{\noindent \sc Note. \ }}

\newcommand{\cws}{{\overset{*}{\rightharpoonup}}}

\newcommand{\epsi}{\varepsilon}
\def\d{{\rm d}}
\def\dx{{\rm d}x}
\def\dy{{\rm d}y}
\def\da{{\rm d}a}
\def\dt{{\rm d}t}
%\def\leq{\leqslant}
%\def\geq{\geqslant}

\def\appendixname{\empty}

\numberwithin{equation}{section}

\newtheoremstyle{thmlemcorr}{10pt}{10pt}{\itshape}{}{\bfseries}{.}{10pt}{{\thmname{#1}\thmnumber{
#2}\thmnote{ (#3)}}}
\newtheoremstyle{thmlemcorr*}{10pt}{10pt}{\itshape}{}{\bfseries}{.}\newline{{\thmname{#1}\thmnumber{
#2}\thmnote{ (#3)}}}
\newtheoremstyle{defi}{10pt}{10pt}{\itshape}{}{\bfseries}{.}{10pt}{{\thmname{#1}\thmnumber{
#2}\thmnote{ (#3)}}}
\newtheoremstyle{remexample}{10pt}{10pt}{}{}{\bfseries}{.}{10pt}{{\thmname{#1}\thmnumber{
#2}\thmnote{ (#3)}}}
\newtheoremstyle{ass}{10pt}{10pt}{}{}{\bfseries}{.}{10pt}{{\thmname{#1}\thmnumber{
A#2}\thmnote{ (#3)}}}

\theoremstyle{thmlemcorr}
\newtheorem{theorem}{Theorem}
\numberwithin{theorem}{section}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{thmlemcorr*}
\newtheorem{theorem*}{Theorem}
\newtheorem{lemma*}[theorem]{Lemma}
\newtheorem{corollary*}[theorem]{Corollary}
\newtheorem{proposition*}[theorem]{Proposition}
\newtheorem{problem*}[theorem]{Problem}
\newtheorem{conjecture*}[theorem]{Conjecture}

\theoremstyle{defi}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{hyp}{Assumption}
\newtheorem{claim}{Claim}
\newtheorem{problem}{Problem}

\theoremstyle{remexample}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem{counterexample}[theorem]{Counterexample}

\newtheorem{teo}[theorem]{Theorem}
\newtheorem{lem}[theorem]{Lemma}
\newtheorem{pro}[theorem]{Proposition}
\newtheorem{cor}[theorem]{Corollary}

%\def\contentsname{}
\def\refname{References}
\def\bibname{References}
\def\chaptername{\empty}
\def\figurename{Fig.}
\def\abstractname{Abstract}

\theoremstyle{ass}

%\theoremstyle{definition}

\newtheorem{exemp}{Exemplo}

\begin{document}

\title[Detection of conserved quantities for PDE schemes]{Algorithmic detection of conserved quantities of finite-difference schemes for partial differential equations}

\author{Diogo A. Gomes}
\affiliation{%
	\institution{King Abdullah University of Science and Technology (KAUST)}
	\department[0]{AMCS/CEMSE Division}
	\city{Thuwal}
	\postcode{23955-6900}
	\country{Saudi Arabia}}
\email{diogo.gomes@kaust.edu.sa}
\author{Friedemann Krannich}
\affiliation{%
	\institution{King Abdullah University of Science and Technology (KAUST)}
	\department[0]{AMCS/CEMSE Division}
	\city{Thuwal}
	\postcode{23955-6900}
	\country{Saudi Arabia}}
\email{friedemann.krannich@kaust.edu.sa}
\author{Ricardo de Lima Ribeiro}
\affiliation{%
	\institution{King Abdullah University of Science and Technology (KAUST)}
	\department[0]{AMCS/CEMSE Division}
	\city{Thuwal}
	\postcode{23955-6900}
	\country{Saudi Arabia}}
\email{ricardo.ribeiro@kaust.edu.sa}

\keywords{Symbolic computations; Finite-difference schemes; Discrete variational derivative; Discrete partial variational derivative; Conserved quantities; Implicit schemes; Explicit schemes.}

\begin{CCSXML}
	<ccs2012>
	<concept>
	<concept_id>10002950.10003714.10003727.10003729</concept_id>
	<concept_desc>Mathematics of computing~Partial differential equations</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	<concept>
	<concept_id>10002950.10003714.10003715.10003750</concept_id>
	<concept_desc>Mathematics of computing~Discretization</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	<concept>
	<concept_id>10002950.10003714.10003715.10003720.10003747</concept_id>
	<concept_desc>Mathematics of computing~Gr{\"o}bner bases and other special bases</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	<concept>
	<concept_id>10010147.10010148.10010149.10010152</concept_id>
	<concept_desc>Computing methodologies~Symbolic calculus algorithms</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	<concept>
	<concept_id>10010147.10010148.10010149.10010155</concept_id>
	<concept_desc>Computing methodologies~Discrete calculus algorithms</concept_desc>
	<concept_significance>500</concept_significance>
	</concept>
	</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Mathematics of computing~Partial differential equations}
\ccsdesc[500]{Mathematics of computing~Discretization}
\ccsdesc[500]{Mathematics of computing~Gr{\"o}bner bases and other special bases}
\ccsdesc[500]{Computing methodologies~Symbolic calculus algorithms}
\ccsdesc[500]{Computing methodologies~Discrete calculus algorithms}

\date{\today}

\begin{abstract}
Many partial differential equations (PDEs) admit conserved quantities like mass or energy. Those quantities are often essential to establish well-posed results. When approximating a PDE by a finite-difference scheme, it is natural to ask whether related discretized quantities remain conserved over time by the scheme. Such conservation may establish the stability of the numerical scheme. We present an algorithm for checking the preservation of a polynomial quantity under a polynomial finite-difference scheme. Our schemes can be explicit or implicit, have higher-order time and space derivatives, and an arbitrary number of variables. 
Additionally, we present an algorithm for, given a scheme, finding conserved quantities. 
We illustrate our algorithm with several finite-difference schemes.
\end{abstract}

\maketitle
\tableofcontents

\begin{acks}
	The authors were supported by \grantsponsor{GN01}{King Abdullah University of Science and Technology (KAUST) baseline funds}{} and 
	\grantsponsor{GN02}{KAUST OSR-CRG2021-4674}{}.
\end{acks}

\section{Introduction}
%\label{intro}
Many partial differential equations (PDEs) admit integral quantities, that are conserved in time. 
For example, the advection equation preserves energy and the heat equation conserves mass. 
When approximating PDEs by a finite-difference scheme, the question arises whether such quantities are conserved by the scheme. 
Such information can be crucial to estimate whether a scheme approximates a PDE accurately and to determine its stability.\\
Computations for determining conservation can get rather tedious.
Hence, automating them in computer algebra systems is desirable.\\
Conserved quantities correspond to conservation laws, admitted by the PDE. A conservation law \cite{cheviakov20} is an equation of the form

@@

	D_t\Phi[u]+D_x\Psi[u]=0

@@

holding for all solutions $u$ of the considered PDE,
that induces the conserved quantity $\int \Phi[u]\ dx$ as
	\frac{d}{dt}\int \Phi[u]\ dx=\int D_t\Phi[u]\ dx=-\int D_x\Psi[u]\ dx=0
assuming periodic boundary conditions in $x$.\\
An analog formulation describes conservation laws for schemes of PDEs, where the derivatives are replaced by shifts \cite{hydon01}.  The key for the construction of schemes, that admit certain conservation laws from the continuous PDE, is the use of the discrete Euler operator \cite{cheviakov20} (also called discrete variational derivative). Kupershmidt \cite[II. Theorem 31]{kupershmidt85} discovered, that an equation is a discrete conservation law if and only if it belongs to the kernel of the discrete Euler operator. Hence, standard approaches for constructing schemes with conservation are either discretizing the continuous conservation law or finding multipliers for the scheme \cite{dorodnitsyn21}, such that the result is in the kernel of the discrete Euler operator.
This strategy was, for example, recently used by Dorodnitsyn et al. to develop a scheme for the shallow water equation, that preserves energy \cite{dorodnitsyn21}. Cheviakov et al. used this idea to find schemes for the linear and nonlinear wave equation, that admit several discrete analogs of continous conservation laws \cite{cheviakov20}.
The same strategies for finding conservation laws, that are described above, work for continuous PDEs or semi-discretized PDEs:
Algorithms for finding conserved quantities, using multipliers and the Euler operator for PDEs were developed and implemented by Cheviakov \cite{cheviakov07,cheviakov10}.
Hereman et al. \cite{hereman04,hereman08}, proposed an algorithm to compute conserved densities for semi-discretized schemes for PDEs with first-order time derivative. 
Their algorithm uses the scaling symmetries of the scheme to construct conserved quantities and calculates their coefficients using the discrete Euler operator.
Gao et al. extended this algorithm to first-order time-explicit schemes \cite{gao02} and first-order time-implicit schemes \cite{gao04}.\\
Bihlo et al. \cite{bihlo12} used another approach and derived finite-difference schemes for the shallow water equations, that are symmetry-invariant on an adaptive mesh and then analyzed them regarding conserved quantities.\\
In this paper, we propose an algorithm that checks if a quantity is conserved in time under a scheme. Our approach differs from the ideas described above, as we do not try to construct discrete $\Phi$ and $\Psi$, using the scheme, but we check if a given discrete $\Phi$ is conserved in time under a given scheme.\\
Conserved quantities usually involve integral expressions or their discrete analog, sums.
Gomes et al. proposed algorithms for the simplification of sums \cite{gomes20}.
They also developed and implemented algorithms for detecting quantities conserved by PDEs and semi-discretized schemes. 
We revise these techniques in Section \ref{assp}. 
To generalize those methods to situations where we do not sum over all arguments of the functions involved, we introduce the discrete partial variational derivative (Section \ref{pvard}).
The main contribution of this paper is an algorithm for checking the conservation of a quantity under a numerical scheme.
Gerdt showed \cite{gerdt12}, that the quantity is conserved, if its discrete time derivative belongs to the difference ideal generated by the scheme. However, some quantities may add to a constant (e.g. telescopic sums) and thus be trivially preserved without belonging to the difference ideal. Moreover, Gerdt's algorithm may not terminate, as the Gr{\"o}bner basis for the difference ideal may not be finite.
We overcome these issues by combining the discrete partial variational derivative with a polynomial ideal (instead of a difference ideal) with finite Gr{\"o}bner basis (Section \ref{algodetec}). 
Our algorithm works for schemes that are explicit and implicit in time and can treat schemes with several higher-order time  and space derivatives, and with several space dimensions.
Additionally, we can handle systems of equations and schemes with parameters. 
We have implemented this algorithm as part of a package in {\sc Mathematica} \cite{wolfram21} \footnote{The code is available upon request. For access contact ricardo.ribeiro@kaust.edu.sa.}.
In the examples, we show that our code finds conserved quantities and proper schemes for the time-implicit and time-explicit discretization of the Burgers equation and a system of PDEs arising in the study of mean-field games (Section \ref{eaa}).

\section{Preliminaries}
\label{assp}
Here, we follow the ideas in \cite{gomes20}.
In this paper, subscripts denote indices related to coordinates or tuples and superscripts denote indices related to sequences.
To avoid boundary terms, we work with periodic functions in $\Zz^d$.

@@

	Let $N$, $d$ and $m$ be positive integers. The \emph{discrete torus} is $\Ii:=\{0,\ldots,N-1\}^d\subset \Zz^d$. Define the space
		\mathcal{P}(\Ii,\Rr^m):=\{u\ |\ u:\Ii\rightarrow\Rr^m\},
	extended to $\Zz^d$ by periodicity.
%Assuming periodicity allows us to extend the functions in $\mathcal{P}(\Ii)$ to $\Zz^d$, and we avoid discussing boundary terms.

@@

In the previous definition, the $e^1,\ldots,e^k$ are not necessarily unit vectors and can vary between functionals.

@@

To simplify sums, the discrete variational derivative is a useful tool.

@@

	Below in \eqref{periodicity}, we shift the indices due to periodicity of $u$ and $v$,

@@

@@

Because we assume all related functions to be smooth, the discrete variational derivative always exists.

@@

	Let us return to Example \ref{functional1} and compute

@@

		=\sum_{n\in\Ii}\mathcal{V}(\mathcal{J})[u](n)v(n)

@@

	and hence $\mathcal{V}(\mathcal{J})[u](n)=-2u(n+1)+4u(n)-2u(n-1)$.

@@

The algorithms for the simplification of sums presented in \cite{gomes20} rely on the following result:
\begin{teo}\label{theo1}
	Let $\mathcal{J},\tilde{\mathcal{J}}\in \mathcal{F}(\mathcal{P}(\Ii,\Rr^m))$. If $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[v]=0$ for all $v\in\mathcal{P}(\Ii,\Rr^m)$ and if there exists $u_0\in\mathcal{P}(\Ii,\Rr^m)$ such that $\mathcal{J}[u_0]=\tilde{\mathcal{J}}[u_0]$, then
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$.
		 $\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u]=0$
	for all $u\in\mathcal{P}(\Ii,\Rr^m)$, if $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{teo}
\begin{proof}
		(\mathcal{J}-\tilde{\mathcal{J}})[u]=(\mathcal{J}-\tilde{\mathcal{J}})[u]-(\mathcal{J}-\tilde{\mathcal{J}})[u_0]\\
		=\int_{0}^{1}\frac{d}{d\lambda}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)]\ d\lambda\\
		=\int_{0}^{1}\mathcal{V}(\mathcal{J}-\tilde{\mathcal{J}})[u_0+\lambda(u-u_0)](u-u_0)\ d\lambda=0
	and hence $\mathcal{J}[u]=\tilde{\mathcal{J}}[u]$.
\end{proof}
We use Theorem \ref{theo1} to examine if different sums represent the same quantity.
	Let $d=m=1$ and consider the functionals
		\mathcal{J}[u]:=\sum_{n\in\Ii}u(n)u(n+1)\text{ and }
		\tilde{\mathcal{J}}[u]:=\sum_{n\in\Ii}u(n-2)u(n-1).
	It is clear, that $\mathcal{J}$ and $\tilde{\mathcal{J}}$ represent the same quantity (by shifting $n$). We confirm this, using the discrete variational derivative:
		D(\mathcal{J}-\tilde{\mathcal{J}})[u](v)=\\
		=\sum_{n\in\Ii}(u(n+1)+u(n-1)-u(n+1)-u(n-1))v(n)=0.

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[latin1]{inputenc}
\usepackage{hyperref}
\usepackage{dsfont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{ifthen}
\usepackage{mmacells}
\usepackage[invisible]{MLaTeX}

\allowdisplaybreaks

\newcommand{\argmin}{\operatorname{argmin}}
\newcommand{\sech}{\operatorname{sech}}
\newcommand{\intr}{\operatorname{int}}
\newcommand{\arccosh}{\operatorname{arccosh}}
\newcommand{\arcsinh}{\operatorname{arcsinh}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\spn}{\operatorname{span}}
\newcommand{\arccotg}{\operatorname{arccotg}}
\newcommand{\inter}{\operatorname{int}}
\newcommand{\var}{\operatorname{Var}}
\newcommand{\supp}{\operatorname{supp}}
\newcommand{\car}{\operatorname{car}}
\newcommand{\cl}{\operatorname{cl}}
\newcommand{\co}{\operatorname{co}}
\newcommand{\arctanh}{\operatorname{arctanh}}
\newcommand{\esssup}{\operatorname{esssup}}
\newcommand{\essinf}{\operatorname{essinf}}
\renewcommand{\div}{\operatorname{div}}
\newcommand{\grad}{\operatorname{grad}}

\newcommand{\Rr}{\mathbb{R}}
\newcommand{\Cc}{\mathbb{C}}
\newcommand{\Nn}{\mathbb{N}}
\newcommand{\Zz}{{\mathbb{Z}}}
\newcommand{\Xx}{\mathbb{X}}
\newcommand{\Qq}{\mathbb{Q}}
\newcommand{\Tt}{\mathbb{T}}
\newcommand{\1}{\chi}

\newcommand{\Hh}{\overline{H}}
\newcommand{\ot}{\overline{t}}
\newcommand{\ox}{\overline{x}}
\newcommand{\oy}{\overline{y}}
\newcommand{\Ll}{\mathcal{L}}

\newcommand{\Aa}{\mathcal{A}}
\newcommand{\Cg}{\mathcal{C}}
\newcommand{\Ff}{\mathcal{F}}
\newcommand{\Gg}{\mathcal{G}}
\newcommand{\Ii}{\mathcal{I}}
\newcommand{\Dd}{\mathcal{D}}
\newcommand{\Uu}{\mathcal{U}}
\newcommand{\Vv}{\mathcal{V}}
\newcommand{\Ww}{\mathcal{W}}
\newcommand{\Mm}{\mathcal{M}}

\newcommand{\eef}{\quad\blacktriangleleft}
\newcommand{\bx}{\boldsymbol{x}}
\newcommand{\bu}{\boldsymbol{u}}
\newcommand{\bdx}{\dot{\boldsymbol{x}}}
\newcommand{\bp}{\boldsymbol{p}}
\newcommand{\bdp}{\dot{\boldsymbol{p}}}
\newcommand{\bX}{\boldsymbol{X}}
\newcommand{\bdX}{\dot{\boldsymbol{X}}}
\newcommand{\bP}{\boldsymbol{P}}
\newcommand{\bdP}{\dot{\boldsymbol{P}}}
\newcommand{\by}{\boldsymbol{y}}
\newcommand{\bdy}{\dot{\boldsymbol{y}}}
\newcommand{\bv}{\boldsymbol{v}}
\newcommand{\bdv}{\dot{\boldsymbol{v}}}
\newcommand{\bw}{\boldsymbol{w}}
\newcommand{\bdw}{\dot{\boldsymbol{w}}}

\newcommand{\cws}{\overset{*}{\rightharpoonup}}

\newcommand{\epsi}{\varepsilon}
\newcommand{\dx}{\mathrm{d}x}
\newcommand{\dy}{\mathrm{d}y}
\newcommand{\da}{\mathrm{d}a}
\newcommand{\dt}{\mathrm{d}t}

\begin{document}
%
\title{Algorithmic detection of conserved quantities for finite-difference schemes\thanks{Supported by King Abdullah University of Science and Technology (KAUST) baseline funds and KAUST OSR-CRG2021-4674.}}
%
%\titlerunning{Abbreviated paper title}
%
\author{Diogo A. Gomes\inst{1}\orcidID{0000-0002-3129-3956} \and
	Friedemann Krannich\inst{1}\orcidID{1111-2222-3333-4444} \and
	Bashayer Majrashi\inst{1}\orcidID{0009-0008-3677-8064 } \and
	Ricardo de Lima Ribeiro\inst{1}\orcidID{0000-0002-2707-2210}}
%
\authorrunning{D.A. Gomes et al.}
%
\institute{King Abdullah University of Science and Technology (KAUST), AMCS/CEMSE Division, Thuwal 23955-6900, Saudi Arabia\
	\email{{diogo.gomes,friedemann.krannich,bashayer.majrashi, ricardo.ribeiro}@kaust.edu.sa}}
%
\maketitle
%
\begin{abstract}
	Many partial differential equations (PDEs) admit conserved quantities like mass or energy. These quantities are often essential for establishing well-posedness results. When approximating a PDE with a finite-difference scheme, it is crucial to determine whether related discretized quantities remain conserved over time by the scheme. Such conservation may ensure the stability of the numerical scheme. We present an algorithm for verifying the preservation of a polynomial quantity under a polynomial finite-difference scheme. Our schemes can be explicit or implicit, have higher-order time and space derivatives, and an arbitrary number of variables. Additionally, we introduce an algorithm for finding conserved quantities, given a scheme. We illustrate our algorithm with several finite-difference schemes. Our approach incorporates parametric G\"obner bases to handle parameters, ensuring accurate computation of conserved quantities.

	\keywords{Symbolic computations \and Finite-difference schemes \and Discrete variational derivative \and Discrete partial variational derivative \and Conserved quantities \and Implicit schemes \and Explicit schemes \and Parametric Groebner basis.}
\end{abstract}

\section{Introduction}
%\label{intro}
In many partial differential equations (PDEs), conserved integral quantities play a crucial role, as they often represent properties of physical significance, such as energy or mass. For example, the advection equation preserves energy and the heat equation conserves mass. 
Since closed-form solutions for these PDEs are often unavailable, finite-difference schemes serve as a practical numerical alternative. Hence, it becomes imperative to determine whether the chosen scheme preserves these conserved quantities. This information is not only valuable for gauging the accuracy of the approximation, but also vital for evaluating its stability. Automating conservation computations in computer algebra systems streamlines the process, saves time, enhances accuracy, and may create new research opportunities.

{\color{red} $\Phi$ and $\Psi$ need domains and ranges, also explain that divergence is only in $x$}

Conserved quantities correspond to conservation laws, admitted by the PDE. A conservation law \cite{cheviakov20} is an equation of the form
\begin{gather*}
	\frac{\partial }{\partial t}\Phi[u]+\div \Psi[u]=0
\end{gather*}
holding for all solutions $u$ of the considered PDE,
that induces the conserved quantity $\int \Phi[u]\ dx$ as
\begin{gather*}
	\frac{d}{dt}\int \Phi[u]\ dx=\int D_t\Phi[u]\ dx=-\int D_x\Psi[u]\ dx=0
\end{gather*}
assuming periodic boundary conditions in $x$.

There is a similar formulation for
for conservation laws for finite difference schemes of a PDE  where the derivatives in the preceding identity are replaced by shifts \cite{hydon01}.  . As a result, standard approaches for constructing schemes with conservation involve either discretizing the continuous conservation law or finding multipliers for the scheme \cite{dorodnitsyn21}.
Alternatively, 
to verify that a particular scheme admits conservation laws, we can use the  discrete Euler operator \cite{cheviakov20}. Kupershmidt \cite[II. Theorem 31]{kupershmidt85} discovered that an equation represents a discrete conservation law if and only if it is in the kernel of the discrete Euler operator
This strategy has been employed in recent studies, such as by Dorodnitsyn et al. for developing a shallow water equation scheme that preserves energy \cite{dorodnitsyn21}, and by Cheviakov et al. for creating linear and nonlinear wave equation schemes that admit discrete analogs of continuous conservation laws \cite{cheviakov20}.
These strategies also apply to continuous or semi-discretized PDEs. Algorithms using multipliers and the Euler operator for PDEs were developed and implemented by Cheviakov \cite{cheviakov07,cheviakov10}. Hereman et al. \cite{hereman04,hereman08} proposed an algorithm for computing conserved densities for semi-discretized PDE schemes with first-order time derivatives, using scaling symmetries and the discrete Euler operator. Gao et al. extended this algorithm to first-order time-explicit \cite{gao02} and time-implicit schemes \cite{gao04}.
Bihlo et al. \cite{bihlo12} used another approach and derived finite-difference schemes for the shallow water equations, that are symmetry-invariant on an adaptive mesh and then analyzed them regarding conserved quantities.

In this paper, we present an algorithm that verifies if a quantity is conserved over time under a given scheme. Our approach diverges from previous methods, as we do not attempt to construct discrete $\Phi$ and $\Psi$ using the scheme, but rather check if a given discrete $\Phi$ is conserved over time under the scheme. Typically, conserved quantities involve integral expressions or their discrete analogs, such as sums. Gomes et al. proposed algorithms for simplifying sums \cite{gomes20} and developed algorithms for detecting conserved quantities in PDEs and semi-discretized schemes \cite{gomes20}. We review these techniques in Section \ref{assp}.
To generalize these methods for cases without summing over all arguments of the involved functions, we introduce the discrete partial variational derivative (Section \ref{pvard}). Our main contribution is an algorithm for verifying the conservation of a quantity under a numerical scheme. Gerdt \cite{gerdt12} demonstrated that the quantity is conserved if its discrete time derivative belongs to the difference ideal generated by the scheme. However, some quantities may add only to a constant and be trivially preserved without being part of the difference ideal. Furthermore, Gerdt's algorithm may not always terminate, as the Gr{\"o}bner basis for the difference ideal might be infinite.
We address these issues by combining the discrete partial variational derivative with a polynomial ideal (instead of a difference ideal) having a finite Gr{\"o}bner basis (Section \ref{algodetec}). Our algorithm is applicable to schemes that are explicit and implicit in time, can handle schemes with multiple higher-order time and space derivatives, multiple space dimensions, systems of equations, and schemes with parameters.
As the process of finding conserved quantities involves polynomial division, special care must be taken to ensure that the division is performed correctly with respect to the parameters. We employ parametric Gr{\"o}bner  bases, which are crucial for the proper treatment of parameters in our algorithm, see Section \ref{pgb}
 We have implemented this algorithm as a part of a {\sc Mathematica} package \cite{wolfram21}\footnote{The code is available upon request. For access, contact ricardo.ribeiro@kaust.edu.sa.}. Our examples include Burgers equation and a PDE system related to mean-field games (Section \ref{eaa}). In these examples, 
 we show that our code detects conserved quantities and can be used to select suitable schemes for time-implicit and time-explicit discretizations.

\section{Preliminaries}
\label{assp}

In this section, we introduce the concept of discrete torus and discrete variational derivatives, and discuss their properties and applications in simplifying sums. 
These definitions are essential for the analysis of methods presented in the rest of the paper.
Here, we 
discuss the connection between discrete functionals and discrete variational derivatives and present examples that illustrate how to simplify sums using this approach.
We follow the ideas in \cite{gomes20}.
In this paper, subscripts denote coordinate or tuple indices, and superscripts denote sequence indices.
To avoid boundary terms, we work with periodic functions in $\Zz^d$.

When discretizing PDEs using finite difference schemes, we need to work with functions defined on discrete sets. For instance, periodic functions in $\Rr^d$ can be replaced by functions defined on the discrete torus, which will be introduced in the following discussion.
	\label{d1}
	Let $N$, $d$ and $m$ be positive integers. The \emph{discrete torus} is the set $\Ii:=\{0,\ldots,N-1\}^d\subset \Zz^d$. Define the space

@@

		\mathcal{P}(\Ii,\Rr^m):=\{u\ |\ u:\Ii\rightarrow\Rr^m\}.

@@

Functions in $\mathcal{P}(\Ii,\Rr^m)$ extend periodically to $\Zz^d$.
We introduce discrete functionals in the previous function space. These functionals are the analog of integral functionals for discrete approximations of functions.

@@

	\label{d2}

@@

In the previous definition, $e^1,\ldots,e^k$ are not necessarily unit vectors and may vary between functionals.

@@

The discrete variational derivative is a useful tool for simplifying sums.

@@

In \eqref{periodicity}, we shift the indices because of the periodicity of $u$ and $v$,

@@

The operator $\mathcal{V}(\mathcal{J})$ can be viewed as a representation of the derivative $D\mathcal{J}u$ with respect to the $L^2$ inner product $\langle w, v\rangle=\sum_{n\in \Ii} w_n v_n$, leading to the following definition.

@@

Because we assume all related functions are smooth, the discrete variational derivative always exists.

@@

Now, we return to Example \ref{functional1} and compute

@@

		=\sum_{n\in\Ii}\mathcal{V}(\mathcal{J})[u](n)v(n).

@@

Thus, $\mathcal{V}(\mathcal{J})[u](n)=-2u(n+1)+4u(n)-2u(n-1)$.

@@

The algorithms for the simplification of sums presented in \cite{gomes20} relies on the following result, which 
provides a criterion to compare functionals and evaluate if they represent the same quantity.
